---
title: "Basic example"
author: "Kyungsub Lee"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Install pacakge

To install ```mHawkes``` package, first install ```devtools```.

```{r, eval=FALSE}
install.packages("devtools")  #if devtools is not installed
```

Install ```mHawkes``` package from github.

```{r, eval=FALSE}
devtools::install_github("ksublee/mHawkes", build_vignettes=FALSE)
```

Load ```mHawkes```.

```{r}
library("mHawkes")
```

### one dimensional Hawkes process

First, create a mHSpec which define the Hawkes model.
Basically the model can be defined up to 9 dimensional 
but currently fully supported for one and two diemsional model especillay for estimation.
The parameter of the model is defined by a matrix but numeric vectors are also supported for one dimesional model.

The following is an example of one dimensional Hawkes model.
Parameter inputs can be a numeric vector or 1 by 1 matrix.
In the following case Jump and eta slots, which deteremine the mark, are default values.

```{r}
MU1 <- 0.3
ALPHA1 <- 1.5
BETA1 <- 2
mHSpec1 <- new("mHSpec", MU=MU1, ALPHA=ALPHA1, BETA=BETA1)
show(mHSpec1)
```

To simulate, use function ```mHSim```.

```{r}
res1 <- mHSim(mHSpec1,  n=10)
```


The output ```res1``` is a list of ```inter_arrival```, ```arrival```, ```jump_type```, ```mark```,```N```, ```Ng```, ```lambda```, and ```lambda_component```.
```inter_arrival```, ```arrival```, ```jump_type```, and ```mark``` are numeric vectors
and ```N```, ```Ng```, ```lambda```, and ```lambda_component``` are matrices.

```{r}
res1
```

Note that the ```inter_arrival``` and ```arrival``` start at zero.
Thus, ```inter_arrival[2]``` and ```arrival[2]``` are first arrival time.


Plot the realized processes.
```{r}
plot(res1$arrival, res1$N[,'N1'], 's', xlab="t", ylab="N")
```
Intensity process can be plotted by ```plotlambda``` function.

```{r}
plotlambda(res1$arrival, res1$N[,'N1'], BETA1)
```

```{r}
inter_arrival1 <- res1$inter_arrival
logLik(mHSpec1, inter_arrival = inter_arrival1)
```



```{r, warning=FALSE}
mle <- mHFit(mHSpec1, inter_arrival = inter_arrival1)
summary(mle)
```


### one dimensional Hawkes process with jump

Mark structure can be added with Jump slot.
In addition, linear impact parameter ```ETA``` can be added.

```{r}
ETA1 <- 0.15
JUMP1 <- function(n,...) rgeom(n, 0.65) + 1

mHSpec1 <- new("mHSpec", MU=MU1, ALPHA=ALPHA1, BETA=BETA1, ETA=ETA1, Jump =JUMP1)
```
```{r}
res1 <- mHSim(mHSpec1,  n=10)
```


Plot the realized processes.

```{r}
plot(res1$arrival, res1$N[,'N1'], 's', xlab="t", ylab="N")
```


### Two-dimensional Hawkes model

```{r}
MU2 <- matrix(c(0.2), nrow = 2)
ALPHA2 <- matrix(c(0.75, 0.92, 0.92, 0.75), nrow = 2, byrow=TRUE)
BETA2 <- matrix(c(2.25, 2.25, 2.25, 2.25), nrow = 2, byrow=TRUE)
ETA2 <- matrix(c(0.19, 0.19, 0.19, 0.19), nrow = 2, byrow=TRUE)
JUMP2 <- function(n,...) rgeom(n, 0.65) + 1

mHSpec2 <- new("mHSpec", MU=MU2, ALPHA=ALPHA2, BETA=BETA2, ETA=ETA2, Jump =JUMP2)
```

```{r}
mHSpec2
```

To simulate, use function ```mHSim```.

```{r}
res2 <- mHSim(mHSpec2,  n=10)
```

Plot N.

```{r}
plot(res2$arrival[1:10], res2$N[1:10,1], 's')
```

```{r}
plot(res2)
```


Plot lambda.

```{r}
plotlambda(res2$arrival[1:10], res2$lambda[1:10,1], BETA2[1,1])
```

Frome the result, we get a vector of realized inter arrival times.
```{r}
inter_arrival1 <- res1$inter_arrival
```
```{r, warning=FALSE}
mle <- mHFit(mHSpec1, inter_arrival = inter_arrival1)
```

Frome the result, we get a vector of realized inter arrival times.
```{r}
inter_arrival2 <- res2$inter_arrival
mark2 <- res2$mark
jump_type2 <- res2$jump_type
```


Log-likelihood is computed by a function ```logLik```.
```{r}
logLik(mHSpec2,  inter_arrival = inter_arrival2, jump_type = jump_type2, mark = mark2)
```

A log-likelihood estimation is performed using ```mHFit```.
```mSpec2``` is regarded as a starting point of the numerical optimization.

```{r, warning=FALSE, error=FALSE}
mle <- mHFit(mHSpec2, inter_arrival = inter_arrival2, jump_type = jump_type2, mark = mark2, constraint=FALSE)
```
```{r}
summary(mle)
```

